---
title: "Environmental parameters - analysis & variable selection"
date: "2018-04-03"
output: 
  html_notebook:
    theme: paper
---
  
Setup!  
```{r setup, include = FALSE}
library(knitr)

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar = c(2, 2, .1, 2))  # smaller margin on top
})

## set the working directory to one up (all notebooks - kept in their own subdirectory within the project directory).
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

## set knitr options for knitting code into the report.
opts_chunk$set(cache = TRUE, # save results so that code blocks aren't re-run unless code changes
               autodep = TRUE, # ..or unless a relevant earlier code block changed
               cache.comments = FALSE, # don't re-run if the only thing that changed was the comments
               highlight = TRUE, 
               small.mar = TRUE)
```

Define the working subdirectories.  
```{r workspace_setup}
## print the working directory, just to be on the safe side
paste("You are here: ", getwd())

data.dir <- "data"    # input data files
functions.dir <- "R"  # functions & scripts
save.dir <- "output"  # clean data, output from models & more complex calculations
figures.dir <- "figs" # plots & figures 
```

Import libraries.  
```{r import_packages, results = FALSE}
library(here) # painless relative paths to subdurectories, etc.
library(tidyverse) # data manipulation, cleaning, aggregation
library(viridis) # smart & pretty colour schemes
library(caret) # variable selection, pruning & other wrapper functions for modeling
library(FactoMineR) # PCA 
library(factoextra) # visualization & exploration of PCA results
library(glmnet) # elastic net variable selection & model fitting 
```

Organize some commonly-used ggplot2 modifications into a more convenient (and less repetitive) format.  
```{r custom_ggplot_settings_helpers}
## ggplot settings & things that I keep reusing
# ggplot_theme <- list(
#   theme_bw(),
#   theme(element_text(family = "Times"))
# )

## always use black-and-white theme
theme_set(theme_bw())

## helper to adjust ggplot text size & avoid repetitions 
text_size <- function(text.x = NULL,
                      text.y = NULL,
                      title.x = NULL,
                      title.y = NULL,
                      legend.text = NULL,
                      legend.title = NULL, 
                      strip.x = NULL, 
                      strip.y = NULL) {
  theme(axis.text.x = element_text(size = text.x),
        axis.text.y = element_text(size = text.y),
        axis.title.x = element_text(size = title.x),
        axis.title.y = element_text(size = title.y),
        legend.text = element_text(size = legend.text), 
        legend.title = element_text(size = legend.title), 
        strip.text.x = element_text(size = strip.x), 
        strip.text.y = element_text(size = strip.y)
        )
}

```

***  

This notebook details the analyses & variable selection of the environmental parameters. I'm giving them their own notebook for clarity and simplicity.  
This is a direct continuation of the cleanup & summarization notebook on environmental parameters; I'm going to import some of the summary tibbles produced there directly from the save folder.    
Again, the analyses will proceed in 3 separate datasets: 2012, sand (2013-2014), and seagrass (2013-2014).   
Two lines of analysis will be applied:  
1) PCA - to see which variables best account for the differences between stations. The "winners" will go on to the multivariate analyses attempting to relate community structure to environment.  
PCA will be applied on 3 sets of explanatory variables - water column (+ LUSI), sediments (+ seagrass, where relevant), and heavy metals; y = station (and habitat, in the case of sand). Prior to the PCA, some (feeble) variable selection will be performed - mostly to get rid of highly correlated variables.  
2) Elastic net - another method of variable selection.  
Again, there will be 3 sets of explanatory variables - as above.  
y = categorical (station, habitats) - logistic regression with multinomial models & grouped lasso penalty.  
y = numeric (species abundances) - multi-response Gaussian linear regression with many correlated responses (+ maybe should be applied over a sparse matrix).  

##### Sand stations (Burgas Bay, 2013-2014)  
Import environmental data.   
```{r import_water_data_sand}
## contains the long-term water column parameters, summarized by station. 
water <- read_csv(here(save.dir, "water_column_summary_LT.csv"))

## filter (manually) to get only the sand 2013-2014 stations
(water.sand <- water %>% 
    filter(station %in% c("Kraimorie", "Chukalya", "Akin", "Sozopol", "Agalina", "Paraskeva"))
)
```

Import the sediment data (it also contains the heavy metals).   
```{r import_sediment_data_sand}
sediments <- read_csv(here(save.dir, "sediments_imputed_sand.csv"))

## split into sediment parameters sesnu stricto, and heavy metals
(sediments.sand <- sediments %>% 
    select(station:year, TOM:silt_clay)
)  

(heavy.metals.sand <- sediments %>% 
    select(station:year, Cu:Ni) 
)
```

Import LUSI and add it to the water column data, although strictly speaking, it isn't really..  
```{r import_lusi_sand}
(lusi <- read_csv(here(save.dir, "lusi_st.csv"))
)

## we'll match them by station - and we only want the sand stations - the rows of x (the water column data) where there are matching values in y (LUSI) -> e.g. inner join 
(water.sand <- inner_join(water.sand, 
                          lusi %>% select(-watershed), 
                          by = "station")
)
```

Now we'll proceed with the variable selection for the PCA. For whatever it's worth, I don't think PCA with so few observations can be valid, but hey, all the kids are doing it.    
We'll just eliminate the most inter-correlated variables using package caret.  
```{r filter_correlated_sand}

# calculate the correlation matrix (on the numeric variables only)
water.sand.cors <- cor(water.sand %>% select(-station))
  
# find and eliminate the highly correlated variables. NB names = FALSE, to return column index
highly.cor.vars.water <- findCorrelation(water.sand.cors, cutoff = 0.85)

ugh <- water.sand %>% 
  select(-station) %>% # otherwise column indices don't match 
  select(-highly.cor.vars.water)


tst.princ <- prcomp(ugh, center = TRUE, scale. = TRUE) 

# check the results
summary(tst.princ)
biplot(tst.princ)

tst.princ$rotation
```

