---
title: "Community structure and composition of macrozoobenthic communities in Sozopol Bay"
subtitle: "Preliminary experiment - 2012"
date: "2018-01-31"
output: 
  html_notebook:
    theme: paper
  
---

Decided to move all 2012 community composition analyses to their own notebook, even if it's kind of clumsy to jump back and forth, because there are way too many workspace objects otherwise.  
This notebook depends heavily on functions, objects & workflow developed for the other, primary experiments. It shouldn't be too much of a problem if it is run **after** the other analyses; I'll make an effort to make it as independent as possible.  
The rest of the biodiversity analyses will be in the corresponding notebooks together with the primary stuff.    

***  
Setup.  
```{r setup, include = FALSE}
library(knitr)
knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar = c(2, 2, .1, 2))  # smaller margin on top
})

## set the working directory to one up (all notebooks - kept in their own subdirectory within the project directory).
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

## set knitr options for knitting code into the report.
opts_chunk$set(cache = TRUE, # save results so that code blocks aren't re-run unless code changes
               autodep = TRUE, # ..or unless a relevant earlier code block changed
               cache.comments = FALSE, # don't re-run if the only thing that changed was the comments
               highlight = TRUE, 
               small.mar = TRUE)
```

Define the working subdirectories.
```{r workspace_setup}
## print the working directory, just to be on the safe side
paste("You are here: ", getwd())

data.dir <- "data"    # input data files
functions.dir <- "R"  # functions & scripts
save.dir <- "output"  # clean data, output from models & more complex calculations
figures.dir <- "figs" # plots & figures 
```
Import the necessary libraries.  
```{r import_packages, message = FALSE}
library(here)
library(vegan)
library(tidyverse)
library(viridis)
```
Import functions.
```{r import_custom_functions}
source(here(functions.dir, "import_data.R"))
```

Organize some commonly-used ggplot2 modifications into a more convenient (and less repetitive) format.
```{r custom_ggplot_settings_helpers}
## ggplot settings & things that I keep reusing
# ggplot_theme <- list(
#   theme_bw(),
#   theme(element_text(family = "Times"))
# )
## always use black-and-white theme
theme_set(theme_bw())

## helper to adjust ggplot text size & avoid repetitions 
text_size <- function(text.x = NULL,
                      text.y = NULL,
                      title.x = NULL,
                      title.y = NULL,
                      legend.text = NULL,
                      legend.title = NULL, 
                      strip.x = NULL, 
                      strip.y = NULL) {
  theme(axis.text.x = element_text(size = text.x),
        axis.text.y = element_text(size = text.y),
        axis.title.x = element_text(size = title.x),
        axis.title.y = element_text(size = title.y),
        legend.text = element_text(size = legend.text), 
        legend.title = element_text(size = legend.title), 
        strip.text.x = element_text(size = strip.x), 
        strip.text.y = element_text(size = strip.y)
        )
}

# log y/min + 1 transform - useful for species counts/biomass data visualization
log_y_min <- function(y) {
  log(y / min(y[y > 0]) + 1)
}
```
***  
#### Macrozoobenthic communities in the seagrass and sand bottoms of Sozopol Bay (2012)  
For this experiment, 3 seagrass sites in Sozpol Bay were chosen (Konski, Ribka, Gradina), with 2 stations within each (total = 6 stations).  
At each station, 4 samples were taken within the seagrass bed, and another 4 - outside it, in the bare sand bottom.  
The purpose was to compare the macrobenthic community composition & structure at the level of the sites, as well as between sites, in the 2 types of habitat (sand and seagrass).   

Import the zoobenthic abundance & biomass data. 
```{r abundance_data_2012_import, message = FALSE}
## import & check the abundance data. Dimensions should be 48 x 205 (5 metadata + 200 species columns); all species columns should be numeric.
(zoo.abnd.2012 <- import_zoo_data(data.dir = data.dir,
                                  zoo.data = "zoo_abnd_2012_orig.csv", 
                                  meta.labels = c("station", "month", "year", "habitat", "replicate")))
```
  
Convert the month and year back to numeric, then arrange the data frame - by station (first converting to factor), then year and month, then habitat. In this experiment, there were 2 habitats - sand and seagrass.
```{r clean_abundance_data_2012}
(zoo.abnd.2012 <- zoo.abnd.2012 %>% 
   mutate(year = as.numeric(year), 
          month = as.numeric(month), 
          station = factor(station, levels = c("Konski1", "Konski2", "Ribka1", "Ribka2", "Gradina1", "Gradina2"))) %>%
   arrange(habitat, station)
)
```
Add the site & arrange.  
```{r clean_abundance_data_2_2012}
(zoo.abnd.2012 <- zoo.abnd.2012 %>% 
   # match 0+ characters at the beginning of the given string which are not digits - because the names of the stations here consist of site + station number.
   mutate(site = str_extract(station, "[^\\d]+")) %>%
   mutate(site = factor(site, levels = c("Konski", "Ribka", "Gradina"))) %>%
  # rearrange to place site at the beginning of the data frame
  select(site, everything()) %>%
  arrange(habitat, site, station)
)
```

Now on to the biomass! 
```{r biomass_data_2012_import, message = FALSE}
## import & check out the biomass data. 
(zoo.biomass.2012 <- import_zoo_data(data.dir = data.dir,
                                     zoo.data = "zoo_biomass_2012_orig.csv",
                                     meta.labels = c("station", "month", "year", "habitat", "replicate")))
# same expected structure as the abundance 
```

```{r clean_biomass_data_2012}
(zoo.biomass.2012 <- zoo.biomass.2012 %>% 
   mutate(year = as.numeric(year),
          month = as.numeric(month), 
          station = factor(station, levels = c("Konski1", "Konski2", "Ribka1", "Ribka2", "Gradina1", "Gradina2"))) %>%
   arrange(habitat, station))
```
  
```{r clean_biomass_data_2_2012}
(zoo.biomass.2012 <- zoo.biomass.2012 %>% 
   # match 0+ characters at the beginning of the given string which are not digits - because the names of the stations here consist of site + station number.
   mutate(site = str_extract(station, "[^\\d]+")) %>%
   mutate(site = factor(site, levels = c("Konski", "Ribka", "Gradina"))) %>%
  # rearrange to place site at the beginning of the data frame
  select(site, everything()) %>%
  arrange(habitat, site, station)
)
```

Great! To be on the safe side, save the cleaned abundance & biomass data frames.
```{r save_clean_data_2012}
write_csv(zoo.abnd.2012, 
          path = here::here(save.dir, "abnd_2012_orig_clean.csv"))
write_csv(zoo.biomass.2012, 
          path = here::here(save.dir, "biomass_2012_orig_clean.csv"))
```
  
Remove the species that are absent from the current dataset. 
```{r clean_data_2012_2}
(zoo.abnd.2012 <- zoo.abnd.2012 %>% 
  gather(key = species, value = sp_count, -c(site:replicate)) %>% 
  group_by(species) %>% 
  filter(sp_count > 0) %>%
  spread(species, sp_count, fill = 0) %>%
  arrange(habitat, site, station))

(zoo.biomass.2012 <- zoo.biomass.2012 %>% 
  gather(key = species, value = sp_biomass, -c(site:replicate)) %>% 
  group_by(species) %>% 
  filter(sp_biomass > 0) %>%
  spread(species, sp_biomass, fill = 0) %>% 
  arrange(habitat, site, station))
```

Transform to long format, too, because will simplify many later analyses.  
```{r long_data_frames_2012}
zoo.abnd.2012.long <- zoo.abnd.2012 %>%
  gather(key = species, value = sp_count, -c(site:replicate))

zoo.biomass.2012.long <- zoo.biomass.2012 %>%
  gather(key = species, value = sp_biomass, -c(site:replicate))
```

Ready to proceed with the community structure & composition analyses! 
  
###### **Summary statistics for the community structure & composition**  
Careful with the multilplication factor when going to ind.m^-2^ & g.m^-2^: all macrobenthic communities in 2012 were sampled with corers!  
The idea of the experiment was to compare seagrass and adjacent bare sand communities; all samples were collected in 2012. Therefore, the two types of habitat will be summarized separately, and will be compared.  

  + the **average abundance** for the **seagrass communities** is `r format(round(mean(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 0), scientific = FALSE)` ind.m^-2^, with standard deviation `r format(round(sd(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 0), scientific = FALSE)`.
  + the **minimum abundance** for the **seagrass communities** is `r format(round(min(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 0), scientific = FALSE)` ind.m^-2^, in sample `r paste(unite(zoo.abnd.2012 %>% filter(row_number() == which.min(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums)), "new", c(station, replicate)) %>% select(new))`.
  + the **maximum abundance** for the **seagrass communities** is `r format(round(max(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 0), scientific = FALSE)` ind.m^-2^, in sample `r paste(unite(zoo.abnd.2012 %>% filter(row_number() == which.max(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums)), "new", c(station, replicate)) %>% select(new))`.
  
  + the **average biomass** for the **seagrass communities** is `r round(mean(filter(zoo.biomass.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 2)` g.m^-2^, with standard deviation `r round(sd(filter(zoo.biomass.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 2)`.
  + the **minimum biomass** for the **seagrass communities** is `r round(min(filter(zoo.biomass.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 2)` g.m^-2^, in sample `r paste(unite(zoo.biomass.2012 %>% filter(row_number() == which.min(filter(zoo.biomass.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums)), "new", c(station, replicate)) %>% select(new))`.
  + the **maximum biomass** for the **seagrass communities** is `r round(max(filter(zoo.biomass.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 2)` g.m^-2^, in sample `r paste(unite(zoo.biomass.2012 %>% filter(row_number() == which.max(filter(zoo.biomass.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% rowSums)), "new", c(station, replicate)) %>% select(new))`.
  
  + the **average abundance** for the **sand communities** is `r format(round(mean(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 0), scientific = FALSE)` ind.m^-2^, with standard deviation `r format(round(sd(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 0), scientific = FALSE)`.
  + the **minimum abundance** for the **sand communities** is `r format(round(min(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 0), scientific = FALSE)` ind.m^-2^, in sample `r paste(unite(zoo.abnd.2012 %>% filter(row_number() == which.min(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums)), "new", c(station, replicate)) %>% select(new))`.
  + the **maximum abundance** for the **sand communities** is `r format(round(max(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 0), scientific = FALSE)` ind.m^-2^, in sample `r paste(unite(zoo.abnd.2012 %>% filter(row_number() == which.max(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums)), "new", c(station, replicate)) %>% select(new))`.
  
  + the **average biomass** for the **sand communities** is `r round(mean(filter(zoo.biomass.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 2)` g.m^-2^, with standard deviation `r round(sd(filter(zoo.biomass.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 2)`.
  + the **minimum biomass** for the **sand communities** is `r round(min(filter(zoo.biomass.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 2)` g.m^-2^, in sample `r paste(unite(zoo.biomass.2012 %>% filter(row_number() == which.min(filter(zoo.biomass.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums)), "new", c(station, replicate)) %>% select(new))`.
  + the **maximum biomass** for the **sand communities** is `r round(max(filter(zoo.biomass.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums) * (1/0.0078), 2)` g.m^-2^, in sample `r paste(unite(zoo.biomass.2012 %>% filter(row_number() == which.max(filter(zoo.biomass.2012, habitat == "S") %>% select(-c(site:replicate)) %>% rowSums)), "new", c(station, replicate)) %>% select(new))`.  
  
  + there are `r ncol(select(zoo.abnd.2012, -c(site:replicate)))` taxa overall. Let's check them out to see which are identifed to species level. Unfortunately, this has to be done by counting manually the ones that are not actually species.
```{r count_number_taxa_2012}
zoo.abnd.2012 %>% 
  select(-c(site:replicate)) %>% 
  names
```
Wow, a whole lot of them are only identified to family level.. Goes to show what I knew about macroinvertebrate taxonomy then, which is zilch. I most likely intended to go back at some point and identify them, but never got round to it, and then destroyed them while measuring the biomass.  
In any case, the 2012 dataset should be regarded as very preliminary, with a really low confidence level, and all conclusions based on it should only be tentative. It will not be included in the other seagrass analyses despite the similar sampling strategy and methods, because due to the low species richness it biases the whole process (e.g. it might be concluded that seagrasses in Sozopol Bay were in a worse ecological state in 2012, which then improved markedly in 2013-2014, but this is not the case - the real variables that improved in that time were my sorting skills and knowledge of taxonomy).   
  
In the **seagrasses** in Sozopol Bay, there were `r zoo.abnd.2012 %>% filter(habitat == "Z") %>% gather(key = species, value = sp_count, -c(site:replicate)) %>% group_by(species) %>% filter(sp_count > 0) %>% count(species) %>% nrow` taxa overall in 2012.  
In the **sands** in Sozopol Bay, there were `r zoo.abnd.2012 %>% filter(habitat == "S") %>% gather(key = species, value = sp_count, -c(site:replicate)) %>% group_by(species) %>% filter(sp_count > 0) %>% count(species) %>% nrow` taxa overall in 2012.  
  
Let's compare the taxa between seagrass and sand communities in Sozopol Bay. 
```{r common_taxa_sand_zostera_2012}
## all taxa found in seagrasses 
taxa.zostera.2012 <- unique(zoo.abnd.2012 %>% 
  filter(habitat == "Z") %>% 
  gather(key = species, value = sp_count, -c(site:replicate)) %>% 
  group_by(species) %>% 
  filter(sp_count > 0) %>% 
  select(species) %>%
  ungroup()
  )

## all taxa found in sand
taxa.sand.2012 <- unique(zoo.abnd.2012 %>% 
  filter(habitat == "S") %>% 
  gather(key = species, value = sp_count, -c(site:replicate)) %>% 
  group_by(species) %>% 
  filter(sp_count > 0) %>% 
  select(species) %>%
  ungroup()
)
```
There are `r length(taxa.zostera.2012$species[which(taxa.sand.2012$species %in% taxa.zostera.2012$species)])` taxa found in both sand and seagrass in 2012. Let's see them:
```{r common_taxa_sand_zostera_2}
# we're looking them up in this order, because there are less taxa in sand than in seagrass in 2012
taxa.zostera.2012$species[which(taxa.sand.2012$species %in% taxa.zostera.2012$species)]
```
The taxa found **only in sand** are:
```{r taxa_sand_only_2012}
setdiff(taxa.sand.2012, taxa.zostera.2012)$species
```
The taxa found **only in seagrass** are: 
```{r taxa_zostera_only_2012}
setdiff(taxa.zostera.2012, taxa.sand.2012)$species
```
  + the **average number of taxa** in **seagrass** is `r round(mean(specnumber(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)))), 0)`, with standard deviation `r round(sd(specnumber(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)))), 0)`.
  + the **minimum number of taxa** in **seagrass** is `r min(specnumber(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate))))`, in sample `r paste(unite(zoo.abnd.2012 %>% filter(row_number() == which.min(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% specnumber)), "new", c(station, replicate)) %>% select(new))`. 
  + the **maximum number of taxa** in **seagrass** is `r max(specnumber(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate))))`, in sample `r paste(unite(zoo.abnd.2012 %>% filter(row_number() == which.max(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)) %>% specnumber)), "new", c(station, replicate)) %>% select(new))`.  
  
  + the **average number of taxa** in **sand** is `r round(mean(specnumber(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)))), 0)`, with standard deviation `r round(sd(specnumber(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)))), 0)`.
  + the **minimum number of taxa** in **sand** is `r min(specnumber(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate))))`, in sample `r paste(unite(zoo.abnd.2012 %>% filter(row_number() == which.min(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)) %>% specnumber)), "new", c(station, replicate)) %>% select(new))`. 
  + the **maximum number of taxa** in **sand** is `r max(specnumber(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate))))`, in sample `r paste(unite(zoo.abnd.2012 %>% filter(row_number() == which.max(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)) %>% specnumber)), "new", c(station, replicate)) %>% select(new))`.  

**Summary statistics by station & habitat**  
  The **total number of taxa per station** in **seagrass** is:
```{r summarize_nb_taxa_2012_by_habitat_st_zostera}
specnumber(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)), groups = filter(zoo.abnd.2012, habitat == "Z")$station)
```
  The **total number of taxa per station** in **sand** is:
```{r summarize_nb_taxa_2012_by_habitat_st_sand}
specnumber(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)), groups = filter(zoo.abnd.2012, habitat == "S")$station)
```

  The **summarized number of taxa** by station and habitat is: 
```{r summarize_nb_taxa_2012_by_station_habitat}
zoo.abnd.2012.long %>%
  group_by(habitat, station, replicate) %>%
  transmute(total_nb_taxa = specnumber(sp_count)) %>%
  distinct(station, habitat, total_nb_taxa) %>%
  group_by(station, habitat) %>% 
  # rounded to the nearest integer, because there is no such thing as half a species 
  summarise(mean_nb_taxa = round(mean(total_nb_taxa)), 
            sd_nb_taxa = round(sd(total_nb_taxa)), 
            nb_replicates = n()) %>%
  arrange(habitat, station)
```

  The **summarized abundance** by station and year is: 
```{r summarize_abundance_2012_by_station_habitat}
## will do it interactively, because it's easier to switch the grouping factors (not so obvious with a function)
zoo.abnd.2012.long %>% 
  group_by(habitat, site, station, replicate) %>%
  transmute(total = sum(sp_count) * 1/0.0078) %>%
  distinct(habitat, station, total) %>%
  group_by(habitat, station) %>%
  summarise(sp_count_mean = round(mean(total), 0), 
            sp_count_sd = round(sd(total), 0), 
            nb_replicates = n())
  

zoo.abnd.2012.long %>% 
  group_by(habitat, site, station, replicate) %>%
  transmute(total = sum(sp_count) * 1/0.0078) %>%
  distinct(habitat, site, total) %>%
  group_by(habitat, site) %>%
  summarise(sp_count_mean = round(mean(total), 0), 
            sp_count_sd = round(sd(total), 0), 
            nb_replicates = n())
  

```
  
  The **summarized biomass** by station and year is: 
```{r summarize_biomass_2012_by_station_habitat}
zoo.biomass.2012.long %>% 
  group_by(habitat, site, station, replicate) %>%
  transmute(total = sum(sp_biomass) * 1/0.0078) %>%
  distinct(habitat, station, total) %>%
  group_by(habitat, station) %>%
  summarise(sp_biomass_mean = round(mean(total), 2), 
            sp_biomass_sd = round(sd(total), 2), 
            nb_replicates = n())

```
<br>  

###### **Taxonomic composition and structure**  
  Leave only the taxa which were present in the 2012 samples (abundance > 0 in at least one sample).
```{r subset_taxonomic_data_2012}
## NB zoo.taxonomy should be imported, if it's not already present from the previous analyses!
zoo.taxonomy <- read_csv(here::here(data.dir, "zoo_taxonomy.csv"))

current.taxa.2012 <- zoo.taxonomy %>%
  filter(species %in% names(zoo.abnd.2012))
```
  
  Add in the more commonly used larger taxonomic groups (Polychaeta, Mollusca, Crustacea, and Varia).
```{r assign_taxonomic_groups_2012}
current.taxa.2012 <- current.taxa.2012 %>%
  mutate(group = case_when(
    phylum == "Mollusca" ~ "Mollusca", 
    phylum == "Annelida" & class == "Polychaeta" ~ "Polychaeta", 
    phylum == "Arthropoda" & subclass %in% c("Eumalacostraca", "Thecostraca") ~ "Crustacea", 
    TRUE ~ "Varia"
  ))
```
  Explore the taxonomic composition of the community: number of taxa per phylum/class, etc.
```{r explore_taxonomic_structure_2012}
## add the taxonomic classification to the 2 subsets of taxa per habitat
taxa.zostera.2012 <- left_join(taxa.zostera.2012, current.taxa.2012)
taxa.sand.2012 <- left_join(taxa.sand.2012, current.taxa.2012)

## number of taxa per class & habitat
# in seagrass
taxa.zostera.2012 %>%
  count(class, sort = TRUE) %>%
  ggplot(mapping = aes(x = reorder(class, n), y = n)) + 
    geom_bar(stat = "identity", fill = "light green") + 
    coord_flip() + 
    labs(title = "Per class - seagrass", x = "", y = "Number of taxa") +
    theme_bw()

# in sand
taxa.sand.2012 %>%
  count(class, sort = TRUE) %>%
  ggplot(mapping = aes(x = reorder(class, n), y = n)) + 
    geom_bar(stat = "identity", fill = "light blue") + 
    coord_flip() + 
    labs(title = "Per class - sand", x = "", y = "Number of taxa") +
    theme_bw()

## number of taxa per phylum & habitat
# in seagrass
taxa.zostera.2012 %>%
  count(phylum, sort = TRUE) %>%
  ggplot(mapping = aes(x = reorder(phylum, n), y = n)) + 
    geom_bar(stat = "identity", fill = "light green") + 
    coord_flip() + 
    labs(title = "Per phylum - seagrass", x = "", y = "Number of taxa") +
    theme_bw()

# in sand
taxa.sand.2012 %>%
  count(phylum, sort = TRUE) %>%
  ggplot(mapping = aes(x = reorder(phylum, n), y = n)) + 
    geom_bar(stat = "identity", fill = "light blue") + 
    coord_flip() + 
    labs(title = "Per phylum - sand", x = "", y = "Number of taxa") +
    theme_bw()
```
  
  Count the number of species by taxonomic group and habitat.
```{r number_taxa_by_tax_group_habitat_2012}
## in seagrass
taxa.zostera.2012 %>%
  count(group)

## in sand
taxa.sand.2012 %>%
  count(group)
```

  Now plot the number of taxa per (common) taxonomic group, and save as file.
```{r plot_number_taxa_by_taxonomic_groups_2012}
nb.taxa.by.gr.habitat.2012 <- left_join(zoo.abnd.2012.long, current.taxa.2012) %>%
  group_by(habitat) %>%
  filter(sp_count > 0) %>%
  distinct(species, group) %>%
  count(group)

(plot.nb.taxa.by.gr.hab.2012 <- ggplot(nb.taxa.by.gr.habitat.2012, mapping = aes(x = reorder(group, n), y = n, fill = habitat)) + 
    geom_bar(stat = "identity", position = "dodge") + 
    scale_fill_manual(values=c("tan1", "light green"), name = "", labels = c("sand", "Zostera")) + 
    coord_flip() + 
    labs(x = "", y = "Number of taxa") +
    theme(axis.text.x = element_text(size = rel(1.3)), 
          axis.text.y = element_text(size = rel(1.3))) +
    theme_bw()
)

ggsave(here::here(figures.dir, "nb_taxa_by_tax_group_2012.png"),
       plot.nb.taxa.by.gr.hab.2012,
       dpi = 300)
```

  In the **seagrass** in Sozopol Bay (2012):  
    + **Crustacea** are represented by `r nrow(taxa.zostera.2012 %>% group_by(group) %>% count(family) %>% filter(group == "Crustacea"))` families.   
    + **Polychaeta** are represented by `r nrow(taxa.zostera.2012 %>% group_by(group) %>% count(family) %>% filter(group == "Polychaeta"))` families.   
    + **Mollusca** are represented by `r nrow(taxa.zostera.2012 %>% group_by(group) %>% count(family) %>% filter(group == "Mollusca"))` families.  
  
  The families with the highest number of species are:
```{r breakdown_nb_taxa_families_2012_zostera}
## in seagrass
lapply(setdiff(unique(taxa.zostera.2012$group), "Varia"), 
       function(x) {
         taxa.zostera.2012 %>%
           group_by(group) %>%
           count(family) %>%
           filter(group == x) %>%
           arrange(desc(n))
       })
```

   In the **sands** in Sozopol Bay (2012):  
   + **Crustacea** are represented by `r nrow(taxa.sand.2012 %>% group_by(group) %>% count(family) %>% filter(group == "Crustacea"))` families.   
   + **Polychaeta** are represented by `r nrow(taxa.sand.2012 %>% group_by(group) %>% count(family) %>% filter(group == "Polychaeta"))` families.   
   + **Mollusca** are represented by `r nrow(taxa.sand.2012 %>% group_by(group) %>% count(family) %>% filter(group == "Mollusca"))` families.  
  
  The families with the highest number of species are:
```{r breakdown_nb_taxa_families_2012_sand}
## in sand
lapply(setdiff(unique(taxa.sand.2012$group), "Varia"), 
       function(x) {
         taxa.sand.2012 %>%
           group_by(group) %>%
           count(family) %>%
           filter(group == x) %>%
           arrange(desc(n))
       })
```
  Now let's calculate the contribution of each taxonomic group to the **overall community composition by abundance**.
```{r tax_group_abnd_contribution_overall_2012}
## helper function to summarize taxonomic group contribution to abundance and biomass
summarize_tax_group_contr <- function(zoo.data, tax.data, key.col, value.col, ...) {
  # zoo.data - long data frame of count/biomass data
  # tax.data - data frame (wide) of taxonomic information on the same species
  # key.col - name of column to be used as key for joining the two data frames
  # value.col - name of column in zoo.data containing the values to be summarized
  
  value.col <- enquo(value.col)
  group.vars <- quos(...) # allow for multiple grouping variables 

  zoo.data %>%
    left_join(tax.data, by = key.col) %>%
    group_by(!!!group.vars) %>%
    summarise(total := sum(!!value.col)) %>%
    mutate(proportion = total / sum(total))
}


zoo.abnd.2012.long %>%
  summarize_tax_group_contr(current.taxa.2012, 
                            key.col = "species", 
                            value.col = sp_count, 
                            habitat, group) %>%
  arrange(habitat, desc(proportion))
```
  Now calculate and plot the **contribution of the taxonomic groups to the abundance by station and habitat** in Sozopol Bay in 2012. 
```{r tax_group_abnd_contrib_st_hab_2012}
tax.props.st.hab.2012 <- zoo.abnd.2012.long %>% 
  summarize_tax_group_contr(current.taxa.2012, 
                            key.col = "species", 
                            value.col = sp_count, 
                            habitat, station, group) %>%
  arrange(habitat, station)

(plot.tax.props.st.hab.2012 <- ggplot(tax.props.st.hab.2012) + 
    geom_bar(mapping = aes(x = station,
                           y = proportion * 100, 
                           fill = group),
             color = "black", size = 0.15, 
             stat = "identity", position = "stack", inherit.aes = TRUE) +
    scale_fill_viridis(option = "viridis", discrete = TRUE, name = "") +
    scale_x_discrete(labels = c("K1", "K2", "R1", "R2", "G1", "G2")) + 
    labs(x = "Station", y = "%") + 
    # change the strip labels
    facet_wrap(~habitat, nrow = 2, labeller = as_labeller(c("S" = "sand", "Z" = "Zostera"))) +
    text_size(text.x = rel(1.3), text.y = rel(1.3), title.x = rel(1.3), legend.text = rel(1.3), strip.x = rel(1.3))
)

## save as file
ggsave(here::here(figures.dir, "tax_gr_props_st_hab_2012.png"),
       plot.tax.props.st.hab.2012,
       dpi = 300)
```

  Check and plot the species which contribute the most to the abundance at each habitat & station.
```{r sp_highest_contrib_abnd_st_2012}
## helper function to get the top n species with the highest abundance/biomass overall
get_top_sp <- function(zoo.data.long, species.col, value.col, nb.sp = 20) {
  # zoo.data - count/biomass data - long format!
  # species.col - name of column containing the species
  # value.col - name of column containing the variable (count, biomass..)
  # nb.sp - number of top-contributing species to select

  # get the columns containing the data
  value_col <- enquo(value.col)
  sp_col <- enquo(species.col)
  
  # find the top n species in the dataset
  top.n <- zoo.data.long %>%
    group_by(!!sp_col) %>%
    summarise(total_count = sum((!!value_col))) %>%
    arrange(desc(total_count)) %>%
    top_n(nb.sp)
  
  return(top.n)
}

top20.sp.2012 <- get_top_sp(zoo.abnd.2012.long, species.col = species, value.col = sp_count, nb.sp = 20)

## subset the species' actual abundance at each station from the original (wide) dataset (to keep the order; I'm too lazy to figure out how to best rearrange the long data itself)
top20.counts.2012.st <- zoo.abnd.2012 %>%
  select(habitat, station, top20.sp.2012$species) %>%
  gather(key = species, value = sp_count, -c(station, habitat)) %>%
  mutate(species = factor(species, levels = unique(species))) # convert species to factor to keep this order for plotting

## plot them
plot.most.abnd.sp.2012 <- ggplot(top20.counts.2012.st, mapping = aes(x = species, y = log_y_min(sp_count), colour = station)) +
  geom_point(size = 2.5, alpha = 0.75) + # make points larger & partially transparent
  scale_color_brewer(palette = "Set2",  labels = c("K1", "K2", "R1", "R2", "G1", "G2")) + 
  ylab("Abundance (log(y/min + 1)") + 
  coord_flip()
  
## some additional tweaks
(plot.most.abnd.sp.2012 <- plot.most.abnd.sp.2012 + 
    facet_wrap(~habitat, ncol = 2, labeller = as_labeller(c("S" = "sand", "Z" = "Zostera"))) +
    text_size(text.x = rel(1.1), title.x = rel(1.1), strip.x = rel(1.1), text.y = rel(1), legend.text = rel(0.9), legend.title = rel(0.9)) +
    theme(legend.position = "top", legend.box.spacing = unit(0.01, "cm"), legend.direction = "horizontal") +
    guides(colour = guide_legend(nrow = 1)) +
    scale_x_discrete(name = "", limits = rev(levels(top20.counts.2012.st$species))) 
)

## save as file
ggsave(here::here(figures.dir, "most_abnd_sp_st_2012.png"), 
       plot.most.abnd.sp.2012, 
       dpi = 300, width = 15, units = "cm")
```

  Now let's calculate the same way the **overall community composition by biomass** in both habitat types in Sozopol Bay (2012).
```{r tax_group_biomass_contrib_overall_2012}
zoo.biomass.2012.long %>% 
  summarize_tax_group_contr(current.taxa.2012, 
                            key.col = "species", 
                            value.col = sp_biomass, 
                            habitat, group) %>%
  arrange(habitat, desc(proportion))
```
  
  Then calculate and plot the  **contribution of the taxonomic groups to the biomass by station and habitat** in Sozopol Bay. 
```{r tax_group_biomass_contribution_st_hab_2012}
tax.props.biomass.st.hab.2012 <- zoo.biomass.2012.long %>% 
  summarize_tax_group_contr(current.taxa.2012, 
                            key.col = "species", 
                            value.col = sp_biomass, 
                            habitat, station, group) %>%
  arrange(habitat, station)

# plot this
(plot.tax.props.biomass.st.hab.2012 <- ggplot(tax.props.biomass.st.hab.2012) + 
    geom_bar(mapping = aes(x = station,
                           y = proportion * 100, 
                           fill = group),
             color = "black", size = 0.15, 
             stat = "identity", position = "stack", inherit.aes = TRUE) +
    scale_fill_viridis(option = "viridis", discrete = TRUE, name = "") +
    scale_x_discrete(labels = c("K1", "K2", "R1", "R2", "G1", "G2")) + 
    labs(x = "Station", y = "%") + 
    # change the strip labels
    facet_wrap(~habitat, nrow = 2, labeller = as_labeller(c("S" = "sand", "Z" = "Zostera"))) +
    text_size(text.x = rel(1.3), text.y = rel(1.3), title.x = rel(1.3), legend.text = rel(1.3), strip.x = rel(1.3))
)

## save as file
ggsave(here::here(figures.dir, "tax_gr_props_biomass_st_hab_2012.png"),
       plot.tax.props.biomass.st.hab.2012,
       dpi = 300)

```

  Combine the proportions of the abundance and the biomass **by station** in one single plot. 
```{r composite_plot_tax_gr_props_st_2012}
## combine the proportional abundance and biomass tibbles into a single common (long) one
tax.props.2012.st.all <- full_join(tax.props.st.hab.2012, 
                                   tax.props.biomass.st.hab.2012,
                                   by = c("habitat", "station", "group")) %>%
  select(-starts_with("total")) %>% 
  gather(key = variable, value = proportion, -habitat, -station, -group) %>%
  ungroup() %>%
  # fix a little the variables that will serve for faceting the plot, to avoid re-labeling later 
  mutate(variable = case_when(variable == "proportion.x" ~ "Abundance", 
                              variable == "proportion.y" ~ "Biomass"), 
         habitat = case_when(habitat == "S" ~ "sand", 
                             habitat == "Z" ~ "Zostera"))


(plot.composite.tax.gr.props.2012.st <- ggplot(tax.props.2012.st.all) + 
    geom_bar(mapping = aes(x = station,
                           y = proportion * 100, 
                           fill = group),
             color = "black", size = 0.15, 
             stat = "identity", position = "stack", inherit.aes = TRUE) +
    scale_fill_viridis(option = "viridis", discrete = TRUE, name = "") +
    scale_x_discrete(labels = c("K1", "K2", "R1", "R2", "G1", "G2")) +
    facet_grid(habitat ~ variable) + 
    labs(x = "Station", y = "%") + 
    theme(legend.position = "top", legend.key.size = unit(0.5, "cm")) +
    text_size(text.x = rel(1.2), text.y = rel(1), title.x = rel(1.2), legend.text = rel(1), strip.x = rel(1.2), strip.y = rel(1.2))
)    
    
ggsave(here::here(figures.dir, "tax_gr_props_2012_st_composite.png"),
       plot.composite.tax.gr.props.2012.st,       
       dpi = 300, width = 15, units = "cm")
```

  Check and plot the species which contribute the most to the biomass at each habitat & station.
```{r sp_highest_contrib_biomass_st_2012}
## calculate the total biomass counts by species, then select the 20 with the highest biomass
top20.sp.biomass.2012 <- get_top_sp(zoo.biomass.2012.long, species.col = species, value.col = sp_biomass, nb.sp = 20)

## subset the species' actual biomass at each station from the original (wide) dataset 
top20.biomass.2012.st <- zoo.biomass.2012 %>%
  select(habitat, station, top20.sp.biomass.2012$species) %>%
  gather(key = species, value = sp_biomass, -c(station, habitat)) %>%
  mutate(species = factor(species, levels = unique(species))) # convert species to factor to keep this order for plotting

## plot them
plot.highest.biomass.sp.st.2012 <- ggplot(top20.biomass.2012.st, mapping = aes(x = species, y = log_y_min(sp_biomass), colour = station)) +
  geom_point(size = 2.5, alpha = 0.75) + # make points larger & partially transparent
  scale_color_brewer(palette = "Set2",  labels = c("K1", "K2", "R1", "R2", "G1", "G2")) + 
  ylab("Biomass (log(y/min + 1)") + 
  coord_flip()
  
## some additional tweaks
(plot.highest.biomass.sp.st.2012 <- plot.highest.biomass.sp.st.2012 + 
    facet_wrap(~habitat, ncol = 2, labeller = as_labeller(c("S" = "sand", "Z" = "Zostera"))) +
    text_size(text.x = rel(1.1), title.x = rel(1.1), strip.x = rel(1.1), text.y = rel(1), legend.text = rel(0.9), legend.title = rel(0.9)) +
    theme(legend.position = "top", legend.box.spacing = unit(0.01, "cm"), legend.direction = "horizontal") +
    guides(colour = guide_legend(nrow = 1)) +
    scale_x_discrete(name = "", limits = rev(levels(top20.biomass.2012.st$species))) 
)

## save as file
ggsave(here::here(figures.dir, "highest_biomass_sp_st_2012.png"), 
       plot.highest.biomass.sp.st.2012, 
       dpi = 300, width = 15, units = "cm")

```
<br>

###### **Site-level comparison of macrobenthic community composition & structure - seagrass/sand, Sozopol Bay, 2012**    

  Check if the community composition differs between sites, in sand/seagrass (using package mvabund for fitting GLMs).
```{r b/n-site_diff_community_composition_2012}
library(mvabund)

## convert the abundance tibble to matrix
zoo.abnd.2012.mat <- mvabund(zoo.abnd.2012 %>% arrange(habitat, site) %>% select(-c(site:replicate)))

## get the sites & habitats
sites.2012 <- zoo.abnd.2012 %>%
  arrange(habitat, site) %>%
  pull(site)

habitats.2012 <- zoo.abnd.2012 %>%
  arrange(habitat, site) %>%
  pull(habitat)

## check the mean-variance relationship in the dataset - for some reason RStudio doesn't want to plot this today - sth to do with graphical device size.
meanvar.plot(zoo.abnd.2012.mat)
```
  Fit the model & check the fit.
```{r b/n-site_diff_2012_model}
## fit the model: do site + habitat influence community composition?
comm.diff.2012.mod <- manyglm(zoo.abnd.2012.mat ~ sites.2012 * habitats.2012, family = "negative.binomial")

## plot resiudals to check model fit
plot(comm.diff.2012.mod)

## check significance of results
comm.diff.2012.mod.summary <- anova(comm.diff.2012.mod, p.uni = "adjusted")
comm.diff.2012.mod.summary
```
  OK, there is a highly significant site, habitat & site:habitat effect on community composition & structure in Sozopol Bay in 2012.
  Also check for differences between stations within each site; if there aren't any, then that's grounds enough to summarize the abundance, biomass,.. at the site level only.
```{r b/n-station_diff_2012_model}
comm.diff.2012.st <- manyglm(zoo.abnd.2012.mat ~ stations.2012 * habitats.2012, famiily = "negative.binomial")

plot(comm.diff.2012.st)

comm.diff.2012.st.summary <- anova(comm.diff.2012.st, p.uni = "adjusted", block = sites.2012)
comm.diff.2012.st.summary
```
  There aren't any statistically significant differences between stations within each site (but there's a significant effect of the station:habitat interaction within each site?). Therefore, we can safely summarize everything appropriate at the **site level**.
  I'll leave the previous station-level summaries in the notebook, too, to show my reasoning, but from now on I'll be summarizing the 2012 data by site.

  NB Save the models & summaries as R data objects, because they take way too much time to repeat.
```{r save_models_2012}
## community difference between sites & habitats
write_rds(comm.diff.2012.mod, here::here(save.dir, "comm_diff_model_2012.RDS"))

write_rds(comm.diff.2012.mod.summary, here::here(save.dir, "comm_diff_model_2012_summary.RDS"))

## community difference between stations & habitats within each site
write_rds(comm.diff.2012.st, here::here(save.dir, "comm_diff_st_model_2012.RDS"))

write_rds(comm.diff.2012.st.summary, here::here(save.dir, "comm_diff_st_model_2012_summary.RDS"))
```

**Summary statistics by site & habitat**  
  The **total number of taxa per site** in **seagrass** is:
```{r summarize_nb_taxa_2012_by_habitat_site_zostera}
specnumber(filter(zoo.abnd.2012, habitat == "Z") %>% select(-c(site:replicate)), groups = filter(zoo.abnd.2012, habitat == "Z")$site)
```
  The **total number of taxa per site** in **sand** is:
```{r summarize_nb_taxa_2012_by_habitat_site_sand}
specnumber(filter(zoo.abnd.2012, habitat == "S") %>% select(-c(site:replicate)), groups = filter(zoo.abnd.2012, habitat == "S")$site)
```

  The **summarized number of taxa** by site and habitat is: 
```{r summarize_nb_taxa_2012_by_site_habitat}
zoo.abnd.2012.long %>%
  group_by(habitat, site, station, replicate) %>%
  transmute(total_nb_taxa = specnumber(sp_count)) %>%
  distinct(site, habitat, total_nb_taxa) %>%
  group_by(site, habitat) %>% 
  # rounded to the nearest integer, because there is no such thing as half a species 
  summarise(mean_nb_taxa = round(mean(total_nb_taxa)), 
            sd_nb_taxa = round(sd(total_nb_taxa)), 
            nb_replicates = n()) %>%
  arrange(habitat, site)
```

  The **summarized abundance** by site and habitat is: 
```{r summarize_abundance_2012_by_site_habitat}
## will do it interactively, because it's easier to switch the grouping factors (not so obvious with a function)
zoo.abnd.2012.long %>% 
  group_by(habitat, site, station, replicate) %>%
  transmute(total = sum(sp_count) * 1/0.0078) %>%
  distinct(site, habitat, total) %>%
  group_by(habitat, site) %>%
  summarise(sp_count_mean = round(mean(total), 0), 
            sp_count_sd = round(sd(total), 0), 
            nb_replicates = n())
  
```
  
  The **summarized biomass** by site and habitat is: 
```{r summarize_biomass_2012_by_site_habitat}
zoo.biomass.2012.long %>% 
  group_by(habitat, site, station, replicate) %>%
  transmute(total = sum(sp_biomass) * 1/0.0078) %>%
  distinct(habitat, site, total) %>%
  group_by(habitat, site) %>%
  summarise(sp_biomass_mean = round(mean(total), 2), 
            sp_biomass_sd = round(sd(total), 2), 
            nb_replicates = n())

```
<br>  

The **contribution of the taxonomic groups to the abundance by site & habitat**:  
```{r tax_group_abnd_contrib_site_hab_2012}
tax.props.site.hab.2012 <- zoo.abnd.2012.long %>% 
  summarize_tax_group_contr(current.taxa.2012, 
                            key.col = "species", 
                            value.col = sp_count, 
                            habitat, site, group) %>%
  arrange(habitat, site)

(plot.tax.props.site.hab.2012 <- ggplot(tax.props.site.hab.2012) + 
    geom_bar(mapping = aes(x = site,
                           y = proportion * 100, 
                           fill = group),
             color = "black", size = 0.15, 
             stat = "identity", position = "stack", inherit.aes = TRUE) +
    scale_fill_viridis(option = "viridis", discrete = TRUE, name = "") +
    # scale_x_discrete(labels = c("Konski", "Ribka", "Gradina")) +
    labs(x = "Site", y = "%") + 
    # change the strip labels
    facet_wrap(~habitat, nrow = 2, labeller = as_labeller(c("S" = "sand", "Z" = "Zostera"))) +
    text_size(text.x = rel(1.3), text.y = rel(1.3), title.x = rel(1.2), legend.text = rel(1), strip.x = rel(1.2))
)

## save as file
ggsave(here::here(figures.dir, "tax_gr_props_site_hab_2012.png"),
       plot.tax.props.site.hab.2012,
       dpi = 300, width = 15, height = 10, units = "cm")
```
  
And summarize the species contributing the most to the abundance at the site level: 
```{r sp_highest_contrib_abnd_site_2012}
## subset the species' actual abundance at each site from the original (wide) dataset
top20.counts.2012.site <- zoo.abnd.2012 %>%
  select(habitat, site, top20.sp.2012$species) %>%
  gather(key = species, value = sp_count, -c(site, habitat)) %>%
  mutate(species = factor(species, levels = unique(species))) # convert species to factor to keep this order for plotting

## plot them
plot.most.abnd.sp.2012.site <- ggplot(top20.counts.2012.site, mapping = aes(x = species, y = log_y_min(sp_count), colour = site)) +
  geom_point(size = 2.5, alpha = 0.75) + # make points larger & partially transparent
  scale_color_brewer(palette = "Set2",  labels = c("Konski", "Ribka", "Gradina")) + 
  ylab("Abundance (log(y/min + 1)") + 
  coord_flip()
  
## some additional tweaks
(plot.most.abnd.sp.2012.site <- plot.most.abnd.sp.2012.site + 
    facet_wrap(~habitat, ncol = 2, labeller = as_labeller(c("S" = "sand", "Z" = "Zostera"))) +
    text_size(text.x = rel(1.1), title.x = rel(1.1), strip.x = rel(1.1), text.y = rel(1), legend.text = rel(0.9), legend.title = rel(0.9)) +
    theme(legend.position = "top", legend.box.spacing = unit(0.01, "cm"), legend.direction = "horizontal") +
    guides(colour = guide_legend(nrow = 1)) +
    scale_x_discrete(name = "", limits = rev(levels(top20.counts.2012.site$species))) 
)

## save as file
ggsave(here::here(figures.dir, "most_abnd_sp_site_2012.png"), 
       plot.most.abnd.sp.2012.site, 
       dpi = 300, width = 15, units = "cm")
```

Plot the **contribution of the taxonomic groups to the biomass by site & habitat**: 
```{r tax_group_biomass_contrib_site_hab_2012}
tax.props.biomass.site.hab.2012 <- zoo.biomass.2012.long %>% 
  summarize_tax_group_contr(current.taxa.2012, 
                            key.col = "species", 
                            value.col = sp_biomass, 
                            habitat, site, group) %>%
  arrange(habitat, site)

(plot.tax.props.biomass.site.hab.2012 <- ggplot(tax.props.biomass.site.hab.2012) + 
    geom_bar(mapping = aes(x = site,
                           y = proportion * 100, 
                           fill = group),
             color = "black", size = 0.15, 
             stat = "identity", position = "stack", inherit.aes = TRUE) +
    scale_fill_viridis(option = "viridis", discrete = TRUE, name = "") +
    # scale_x_discrete(labels = c("Konski", "Ribka", "Gradina")) +
    labs(x = "Site", y = "%") + 
    # change the strip labels
    facet_wrap(~habitat, nrow = 2, labeller = as_labeller(c("S" = "sand", "Z" = "Zostera"))) +
    text_size(text.x = rel(1.3), text.y = rel(1.3), title.x = rel(1.2), legend.text = rel(1), strip.x = rel(1.2))
)

## save as file
ggsave(here::here(figures.dir, "tax_gr_props_biomass_site_hab_2012.png"),
       plot.tax.props.biomass.site.hab.2012,
       dpi = 300, width = 15, height = 10, units = "cm")

```

Composite taxonomic group proportion in the abundance and biomass structure of the **sites**:  
```{r composite_plot_tax_gr_props_site_2012}
## combine the proportional abundance and biomass tibbles into a single common (long) one
tax.props.2012.site.all <- full_join(tax.props.site.hab.2012, 
                                     tax.props.biomass.site.hab.2012,
                                     by = c("habitat", "site", "group")) %>%
  select(-starts_with("total")) %>% 
  gather(key = variable, value = proportion, -habitat, -site, -group) %>%
  ungroup() %>%
  # fix a little the variables that will serve for faceting the plot, to avoid re-labeling later 
  mutate(variable = case_when(variable == "proportion.x" ~ "Abundance", 
                              variable == "proportion.y" ~ "Biomass"), 
         habitat = case_when(habitat == "S" ~ "sand", 
                             habitat == "Z" ~ "Zostera"))


(plot.composite.tax.gr.props.2012.site <- ggplot(tax.props.2012.site.all) + 
    geom_bar(mapping = aes(x = site,
                           y = proportion * 100, 
                           fill = group),
             color = "black", size = 0.15, 
             stat = "identity", position = "stack", inherit.aes = TRUE) +
    scale_fill_viridis(option = "viridis", discrete = TRUE, name = "") +
    facet_grid(habitat ~ variable) + 
    labs(x = "Site", y = "%") + 
    theme(legend.position = "top", legend.key.size = unit(0.5, "cm")) +
    text_size(text.x = rel(1.2), text.y = rel(1), title.x = rel(1.2), legend.text = rel(1), strip.x = rel(1.2), strip.y = rel(1.2))
)    
    
ggsave(here::here(figures.dir, "tax_gr_props_2012_site_composite.png"),
       plot.composite.tax.gr.props.2012.site,       
       dpi = 300, width = 15, height = 10, units = "cm")
```

Plot the species contributing the most to the biomass at the **site** level: 
```{r sp_highest_contrib_biomass_site_2012}
## subset the species' actual biomass at each site from the original (wide) dataset
top20.biomass.2012.site <- zoo.biomass.2012 %>%
  select(habitat, site, top20.sp.biomass.2012$species) %>%
  gather(key = species, value = sp_biomass, -c(site, habitat)) %>%
  mutate(species = factor(species, levels = unique(species))) # convert species to factor to keep this order for plotting

## plot them
plot.highest.biomass.sp.2012.site <- ggplot(top20.biomass.2012.site, mapping = aes(x = species, y = log_y_min(sp_biomass), colour = site)) +
  geom_point(size = 2.5, alpha = 0.75) + # make points larger & partially transparent
  scale_color_brewer(palette = "Set2",  labels = c("Konski", "Ribka", "Gradina")) + 
  ylab("Biomass (log(y/min + 1)") + 
  coord_flip()
  
## some additional tweaks
(plot.highest.biomass.sp.2012.site <- plot.highest.biomass.sp.2012.site + 
    facet_wrap(~habitat, ncol = 2, labeller = as_labeller(c("S" = "sand", "Z" = "Zostera"))) +
    text_size(text.x = rel(1.1), title.x = rel(1.1), strip.x = rel(1.1), text.y = rel(1), legend.text = rel(0.9), legend.title = rel(0.9)) +
    theme(legend.position = "top", legend.box.spacing = unit(0.01, "cm"), legend.direction = "horizontal") +
    guides(colour = guide_legend(nrow = 1)) +
    scale_x_discrete(name = "", limits = rev(levels(top20.biomass.2012.site$species))) 
)

## save as file
ggsave(here::here(figures.dir, "highest_biomass_sp_site_2012.png"), 
       plot.highest.biomass.sp.2012.site, 
       dpi = 300, width = 15, units = "cm")
```
