---
title: "Black Sea currents"
date: "2018-06-27"
output: html_notebook
theme: paper
---

##  Black Sea circulation  
This notebook details the process of opening a NetCDF dataset downloaded from CMEMS (marine.copernicus.eu), extracting & then plotting the data.  

The current dataset contains a subset of surface current velocities from BLKSEA_REANALYSIS_PHYS_007_004 - a reanalysis product of the physical state of the Black Sea (1992-2017). The subset covers the SW Black Sea - Burgas Bay to the Turkish border, and the period 2012-01-01 to 2015-01-01 (my PhD study period).  

***  

Setup!  
```{r setup, include = FALSE}
library(knitr)

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar = c(2, 2, .1, 2))  # smaller margin on top
})

## set the working directory to one directory up (I'm keeping the notebooks in their own subdirectory of the project, doc).
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

## set knitr options for knitting code into the report.
opts_chunk$set(cache = TRUE, # save results so that code blocks aren't re-run unless code changes
               autodep = TRUE, # ..or unless a relevant earlier code block changed
               cache.comments = FALSE, # don't re-run if the only thing that changed was the comments
               highlight = TRUE, 
               small.mar = TRUE)

```

Define working subdirectories. 
```{r workspace_setup}
## print the working directory, just to be on the safe side
paste("You are here: ", getwd())

data.dir <- "data"    # input data files
functions.dir <- "R"  # functions & scripts
save.dir <- "output"  # clean data, output from models & more complex calculations
figures.dir <- "figs" # plots & figures 
```

Import libraries.
```{r import_packages, message = FALSE}
library(here) ## find root of project directory
library(tidyverse) ## data manipulation & tidying
#library(viridis) ## pretty & readable colour schemes 

library(ncdf4) ## read & manipulate NetCDF files
```

Some commonly-used ggplot2 modifications..  
```{r custom_ggplot_settings_helpers}
## ggplot settings & things that I keep reusing
# ggplot_theme <- list(
#   theme_bw(),
#   theme(element_text(family = "Times"))
# )

## always use black-and-white theme
theme_set(theme_bw())

## helper to adjust ggplot text size & avoid repetitions 
text_size <- function(text.x = NULL,
                      text.y = NULL,
                      title.x = NULL,
                      title.y = NULL,
                      legend.text = NULL,
                      legend.title = NULL, 
                      strip.x = NULL, 
                      strip.y = NULL) {
  theme(axis.text.x = element_text(size = text.x),
        axis.text.y = element_text(size = text.y),
        axis.title.x = element_text(size = title.x),
        axis.title.y = element_text(size = title.y),
        legend.text = element_text(size = legend.text), 
        legend.title = element_text(size = legend.title), 
        strip.text.x = element_text(size = strip.x), 
        strip.text.y = element_text(size = strip.y)
        )
  }
```

***  
First, the **daily means** of the current velocities.  
Read in the data & check the file contents.  
```{r import_netcdf}
(currents.in <- nc_open(file = here(data.dir, "currents", "sv04-bs-cmcc-cur-rean-d_1530092744997.nc"))
)
```

Looks fine! Let's extract the current velocities..  
```{r extract_variables}
## meridional current matrix
currents.y <- ncvar_get(currents.in, "vomecrty")
dim(currents.y)

## zonal current matrix
currents.x <- ncvar_get(currents.in, "vozocrtx")
dim(currents.x)

```
The current velocities are stored in a lon x lat x time matrix. There is also a depth dimension, but it only has one value, 2.5 m - only surface currents here.  
Let's extract the time, latitude and longitude - stored as **NetCDF dimensions**.  
```{r extract_dimensions}
## check the names of the dimensions 
attributes(currents.in$dim)$names

## we'll only get the latitude, longitude & time
nc_time <- ncvar_get(currents.in, "time")
nc_lat <- ncvar_get(currents.in, "lat")
nc_lon <- ncvar_get(currents.in, "lon")

## print the dimensions to see if they match what we have for the velocities above..
print(paste(dim(nc_lon), "longitudes, ", dim(nc_lat), "latitudes and ", dim(nc_time), "times."))
```

Reformat the time into something more readable (otherwise it's in seconds since 1990-01-01, as the attributes suggest).  
```{r format_time}
nc_time <- as.POSIXct(nc_time, format = "%Y-%m-%d", tz = "UTC", origin = "1990-01-01")

head(nc_time)
```

Great. Let's get some of the additional data that might come in handy, then close the NetCDF connection. 



```{r}
ugh <- currents.x
dimnames(ugh) <- list(lon = nc_lon, lat = nc_lat, time = nc_time)
as.data.frame.table(ugh, responseName = "value") 

plyr::adply(ugh, 1:3)

```

